---
output:
  html_document:
    theme: readable
    highlight: tango
    self_contained: false
    css: textbook.css
---

# Summarizing Vectors

<br>
<br>

<div class="tip">

## Key Concepts

In this chapter, we'll explore the following key concepts:

* Counting Logical Values: Function `sum()`
* Tallying Categorical Variables: `table()`
* Determining Missing Values: `is.na()`
* Table Proportionality: `prop.table()`
* Table Margins: `margin.table()`
* Functions for Descriptive Stats: 
    - `min()`
    - `max()`
    - `mean()`
    - `summary()`
    - `quantile()`
* Combining Functions `summary()` & `data.frame()`
* Creating Tables of Descriptives: Factor & Numeric Variables

## New Packages

This chapter uses the following packages:

* Package A

## Key Takeaways

Too long; didn't read? Here's what you need to know:

* Item A
* Item B
* Item C

<br>
<br>
<br>

</div>

```{r echo=F}

# ATTENTION : GLOBAL CHUNK DEFAULTS

knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE)

```

```{r include=F}

tutorial::go_interactive( greedy=FALSE )

```

<br>
<br>

## Logical Variables

Class **logical** variables are comprised of one of two values: `TRUE` or `FALSE`.

<br>

We can create **logical** variables manually. Determine their class with function `class()`:

```{r}

x <- TRUE
y <- FALSE

class(x)
class(y)

```

<br>

Often, **logical** values simply indicate "yes" or "no"

```{r}

character <- c("Raskolnikov", "Marmeladov", "Razumikhin")
criminal <- c(TRUE, FALSE, FALSE)

data.frame(character,
           criminal)        # Create data frame of characters and crime status

```

<br>

Most common, however, is that **logical** values result from conditional statements.

Observe the function call `head()` on *ggplot2* dataset `economics`:

```{r}

library(ggplot2)

head(economics)             # First 6 rows of "economics" dataset

```

<br>

Here we use the operator `>` to determine months with more than 10 M unemployed:

```{r}

economics$unemploy > 10000  # Condition: Is unemployment greater than 10 M?

```

<br>

Here's a more complex conditional statement: More than 5 M but less than 10 M:

```{r}

economics$unemploy > 5000 & economics$unemploy < 10000

```
